Predicting political party affiliation from text

### 2016-06-17 manuscript/poltext.tex
Shortened abstract, formatted authors.

\author{Felix Biessmann\\
%  {\tt felix.biessmann@gmail.com} \\ 
  \And
  Sebastian Schelter\\
  Technische Universit\"at Berlin\\
%  {\tt sebastian.schelter@tu-berlin.de}\\
\And 
Daniel Kirsch
  \And
 Pola Lehmann \\
  Wissenschaftszentrum Berlin \\
%  {\tt pola.lehmann@wzb.eu} \\
  }

@@

\begin{abstract}
Every day large amounts of text are produced in public discourse. Some of this text is produced by actors whose political colour is very obvious, for example in the case of party elites. But there are also many actors raising their voices in the discourse, who cannot be clearly attributed to a political party, while their statements might actually by biased towards a specific party. Identifying such biases is crucial for political research as well as media consumers, especially when analysing the influence of the media on political discourse and vice versa. In this study we investigate to what extent political party affiliation can be predicted from textual content. Results indicate that automatic classification of political affiliation is possible with above chance accuracy also across text domains. We propose methods to better interpret these results and find that features not related to political policies, such as speech sentiment, can be discriminative and thus exploited by text analysis models. 
%We present an example of a web application that combines topic modelling with automatic political affiliation analyses of news articles based on the proposed model.
\end{abstract}

@@

The following section gives an overview of the results for all political bias prediction tasks. 
%Some interpretations of the results are highlighted and a web application of the models is presented at the end of the section. 
Due to space restrictions we only report results from the 17th legislative period. Results are similar for the 18th legislative period.

@@

\begin{table}[t]
\caption{
\label{tab:results_topic}
{\bf Out-of-domain} classification performance (topic level) on {\bf manifesto data} of a classifier trained on the 17th legislative period of the Bundestag. Compared to quasi-sentence level predictions (\autoref{tab:results_out-of-domain}) the predictions made on topic level are more reliable.}
\begin{center}
\begin{tabular}{lcccc}
    &         precision    &recall &  f1-score  & N  \\
    \hline
        \hline
cducsu     &  0.64  &    1.00  &    0.78    &     7\\
       fdp    &   1.00    &  1.00    &  1.00    &     7\\
    gruene  &     1.00  &    0.86  &    0.92    &     7\\
     linke    &   1.00   &   1.00     & 1.00    &     7\\
       spd   &    0.80   &   0.50    &  0.62     &    8\\
    \hline
total  &     0.88   &   0.86   &   0.86  &      36\\
\end{tabular}
\end{center}

@@

% Confusion matrix topic level
\begin{table}[t]\label{tab:conf_mat_four_class}
\caption{\label{tab:confusion_topic} {\bf Topic level confusion matrices} on evaluation data (manifesto texts), classifiers were trained on speeches of 17th Bundestag.}
\vspace{0.5em}
\begin{tabular}{lc|ccccc}
&& \multicolumn{5}{c}{\bf Predicted}\\
&& cducsu & fdp& gruene& linke& spd\\
\hline
\multirow{5}{*}{\rotatebox{90}{\pbox{2cm}{\centering {\bf True}}}} &cducsu &7& 0& 0& 0& 0\\
&fdp&0& 7& 0& 0& 0\\
&gruene&0& 0& 6& 0& 1\\
&linke&0& 0& 0& 7& 0\\
&spd&4& 0& 0& 0& 4\\
\end{tabular}
\end{table}

@@

%\subsection{An example web application}
%To show an example use case of the above models a web application was implemented that downloads regularly all articles from some major German news paper websites\footnote{\url{http://www.spiegel.de/politik}, \url{http://www.faz.net/aktuell/politik}, \url{http://www.welt.de/politik}, \url{http://www.sueddeutsche.de/politik}, \url{http://www.zeit.de/politik}} and applies some simple topic modelling to them. For each news article topic, headlines of articles are plotted along with the predictions of the policy issue of an article and two labels derived deterministically from the 56 class output, a left right index and the political domain of a text, see \cite{leftright}. Within each topic it is then possible to get an ordered (from left to right) overview of the articles on that topic. A preliminary demo is live at \cite{fipidemo} and the code is available on github\cite{fipi}.
%\begin{figure}
%\begin{center}
%\includegraphics[width=10cm]{images/fipi-screenshot}
%%
%\end{center}
%\caption{
%\label{fig:fipi}
%A screen shot of an example web application using the political view prediction combined with topic modelling to provide a heterogeneous overview of a topic. }
%\end{figure}

\section{Conclusions and Limitations}\label{sec:conclusion}
Evaluating classifiers trained on parliament speeches we find that automated political bias prediction is possible with above chance accuracy in some cases even beyond the training text domain. These results encourage to use such systems as assistive technology for instance for human annotators in an active learning setting.

>>>>

\author{Felix Biessmann\thanks{~\tt felix.biessmann@gmail.com}\\
  \And
  Sebastian Schelter \thanks{~\tt sebastian.schelter@tu-berlin.de}\\ Technische Universit\"at Berlin
\And 
Daniel Kirsch
  \And
 Pola Lehmann\thanks{ ~{\tt pola.lehmann@wzb.eu} }\\
  Wissenschaftszentrum Berlin\\
}

@@

\begin{abstract}
Every day large amounts of text are produced in public discourse. Some of this text is produced by actors whose political colour is very obvious. But there are also many actors, who cannot be clearly attributed to a political party, while their statements might be biased towards a specific party. Identifying such biases is crucial for political research as well as media consumers, especially when analysing the influence of the media on political discourse and vice versa. In this study we investigate to what extent political party affiliation can be predicted from textual content. Results indicate that automatic classification of political affiliation is possible with above chance accuracy also across text domains. We propose methods to better interpret these results and find that features not related to political policies, such as speech sentiment, can be discriminative and thus exploited by text analysis models. 
%We present an example of a web application that combines topic modelling with automatic political affiliation analyses of news articles based on the proposed model.
\end{abstract}

@@

The following section gives an overview of the results for all political bias prediction tasks. 
Predictions compared with the manifesto data were done using models trained on texts from the 17th Bundestag, predictions obtained for facebook post texts were computed with models trained on the 18th Bundestag.

@@

\begin{table}[t]
\caption{
\label{tab:results_topic}
{\bf Out-of-domain} classification performance (topic level) on {\bf manifesto data}. Compared to quasi-sentence level predictions (\autoref{tab:results_out-of-domain}) the predictions made on topic level are more reliable.}
\begin{center}
\begin{tabular}{lcccc}
    &         precision    &recall &  f1-score  & N  \\
    \hline
        \hline
cducsu     &  0.64  &    1.00  &    0.78    &     7\\
       fdp    &   1.00    &  1.00    &  1.00    &     7\\
    gruene  &     1.00  &    0.86  &    0.92    &     7\\
     linke    &   1.00   &   1.00     & 1.00    &     7\\
       spd   &    0.80   &   0.50    &  0.62     &    8\\
    \hline
total  &     0.88   &   0.86   &   0.86  &      36\\
\end{tabular}
\end{center}

@@

% Confusion matrix topic level
\begin{table}[t]\label{tab:conf_mat_four_class}
\caption{\label{tab:confusion_topic} {\bf Topic level confusion matrices} of manifesto texts.}
\vspace{0.5em}
\begin{tabular}{lc|ccccc}
&& \multicolumn{5}{c}{\bf Predicted}\\
&& cducsu & fdp& gruene& linke& spd\\
\hline
\multirow{5}{*}{\rotatebox{90}{\pbox{2cm}{\centering {\bf True}}}} &cducsu &7& 0& 0& 0& 0\\
&fdp&0& 7& 0& 0& 0\\
&gruene&0& 0& 6& 0& 1\\
&linke&0& 0& 0& 7& 0\\
&spd&4& 0& 0& 0& 4\\
\end{tabular}
\end{table}

@@

\section{Conclusions and Limitations}\label{sec:conclusion}
Evaluating classifiers trained on parliament speeches we find that automated political bias prediction is possible with above chance accuracy even beyond the training text domain. These results encourage to use such systems as assistive technology for instance for human annotators in an active learning setting.
